Modelling the travel time of transit vehicles in \rt through a GTFS-based road network using GPS vehicle locations

### 2019-08-08 anzjs2018/sections/01_introduction.tex
more edits

A more robust modelling and prediction framework 
based on \rt congestion information would avoid making these assumptions.
Such a framework should consist of a robust vehicle model to estimate the position and speed
of transit vehicles from \rt GPS data,
and secondly a means of combining speed information from vehicles
to model traffic flows,
which can then be used to improve arrival time predictions.
Several vehicle modelling approaches were explored, 
including the Kalman filter \citep{Dailey_2001,Cathey_2003},
machine learning models \citep{Yu_2006,Chang_2010},
and the particle filter,
which has successfully been used in vehicle tracking applications
\citep{cn},
as well as in transport modelling \citep{Hans_2015}.
For several reasons, discussed further in Section~\ref{sec:pf},
the particle filter was used in the current application.s

>>>>

A more robust modelling and prediction framework 
based on \rt congestion information would avoid making these assumptions.
Such a framework should consist of a robust vehicle model to estimate the position and speed
of transit vehicles from \rt GPS data,
and secondly a means of combining speed information from vehicles
to model traffic flows,
which can then be used to improve arrival time predictions.
Several vehicle modelling approaches were explored, 
including the Kalman filter \citep{Dailey_2001,Cathey_2003},
machine learning models \citep{Yu_2006,Chang_2010},
and the particle filter \citep{Hans_2015}.

The particle filter has proven itself as a robust option for
\rt vehicle tracking applications
\citep{Gustafsson_2002,Davidson_2011},
and as computational power has increased over recent years,
a strong competitor for \rt applications.
\cite{Ulmke_2006} showed it to handle situations where traditional
(i.e., Kalman filter) methods break down:
the particle filter retained the accurate uncertainty about the location
where there was a lack of information,
an important property for our model in which we expect
many situations where there are sparse observations
and a strongly multimodal distribution.
By using a particle filter,
we were able to develop a simple, flexible framework
which is robust to the type of data we expect to observe.

### 2019-08-08 anzjs2018/sections/02_gtfs.tex
more edits

As mentioned in Section~\ref{sec:intro},
there are some major issues with the arrival time prediction method currently
deployed in Auckland,
which is based entirely on \emph{GTFS-realtime} trip updates---%
vehicle positions are used only to display \emph{where} the bus is.
Trip updates are reported whenever a vehicle arrives at or departs from a stop,
and includes the delay between the scheduled and actual arrival times,
which is propagated to all future stops to update their ETAs.
As already mentioned, this assumes that the schedule is well calibrated,
and that the time between stops
is representative of the real-world travel time between them. 
Figure~\ref{fig:gtfs-delays} demonstrates the fallibility of the current approach,
where we see the progressive lateness causing the passengers' ETA to 
decrease much slower than expected.
An estimation approach using \rt travel times 
between stops would seem more adequate.

>>>>

As mentioned in Section~\ref{sec:intro},
there are some major issues with the arrival time prediction method currently
deployed in Auckland,
which is based entirely on \emph{GTFS-realtime} trip updates---%
vehicle positions are used only to display \emph{where} the bus is.
Trip updates are reported whenever a vehicle arrives at or departs from a stop,
and includes the delay between the scheduled and actual arrival times,
which is propagated to all future stops to update their ETAs.
As already mentioned, this assumes that the schedule is well calibrated,
and that the time between stops
is representative of the real-world travel time between them. 
Figure~\ref{fig:gtfs-delays} demonstrates the fallibility of the current approach,
where we see the progressive lateness causing the passenger's ETA to 
jump each time the bus arrives at a stop.
An estimation approach using \rt travel times 
between stops would seem more adequate.

\begin{figure}[tb]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/02_gtfs_delays.pdf}
    \caption{
        On the left, scheduled arrival times are shown as solid points,
        for several stops along a route. The actual arrival times
        are shown with crosses,
        and the GTFS-based ETAs for stop 5, $\hat A_1$, $\hat A_2$, and $\hat A_3$, are also shown.
        On the right,
        for a passenger arriving at stop 5 at 12:04,
        the ETA over time is shown, demonstrating the fluctuation in
        arrival time as the bus arrives at each stop.
    }
    \label{fig:gtfs-delays}
\end{figure}

### 2019-08-08 anzjs2018/sections/03_models.tex
more edits

The underlying vehicle state at time $t_k$ consists of
the vehicle's distance travelled $x$ in meters along the route and
its speed $\dot x$ in meters per second.
These are combined into the state vector
$\bX_k = \left[x_k\ \dot x_k\right]^\top$,
of which observations $\bY_k$ are made using a GPS
with GPS error $\omega^2$,
giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle
as $\bY_k = \left[\lambda_k\ \phi_k \right]^\top$.

We use a particle filter to estimate $\bX_k$,
due to its high flexibility and use in recent 
transit vehicle modelling applications \citep{Hans_2015},
as well as allowing estimation of travel times
$\bz = (z_1,\ldots,z_L)^\top$, in seconds, along road segments
as the vehicle traverses the route.
The primary advantage of the particle filter is its handling of multimodality,
as demonstrated in Figure~\ref{fig:pf_state_predict},
which is a common feature of the proposal distribution, particularly around bus stops.
Another advantage is that the likelihood function is intuitively based
on the physical distance between the vehicle observation and the state's
estimate of the true location, $h(\bX_k)$ (Section~\ref{sec:pf_update}).
Conversely, particle filter methods are computationally demanding,
requiring an increasing number of particles as model complexity and
number of parameters increases \citep{Carpenter_1999}.
Section~\ref{sec:rt} describes our implementation of the particle filter
and its \rt performance.

@@

The combination of (\ref{eq:trans_newton}) and (\ref{eq:trans_stop})
is handled by the transition function algorithm,
displayed in Algorithm~\ref{fig:algorithm}.
While (\ref{eq:trans_stop}) is used directly in the algorithm,
(\ref{eq:trans_newton}) is evaluated iteratively,
allowing the bus to approach a stop
and conditionally wait before continuing,
until $\delta_k$ reaches zero.
The algorithm also keeps track of each particle's travel time $z_\ell^{(i)}$
along each segment $\ell$ of the route,
storing the times so that the posterior distribution of the travel time
can be computed once all particles have completed segment $\ell$:
\begin{equation}
\label{eq:post_tt}
p(z_\ell | \bY_{1:k}) \approx
    \sum_{i=1}^N W_k^{(i)} \dot\delta_{z_\ell}(z_\ell^{(i)}).
\end{equation}

\renewcommand{\algorithmicrequire}{\textbf{Start:}}
\newcommand{\algorithmicbreak}{\textbf{break}}
\newcommand{\Break}{\State \algorithmicbreak}
\begin{algorithm}[t]
    \caption{Particle mutation function.}
    \label{fig:algorithm}
    \begin{algorithmic}[2]
    \Require $x_{k-1}, \dot x_{k-1}, \boldsymbol z$\\
    Initialisation $\delta \gets t_k - t_{k-1}$, 
        $j \gets stop\_index(x, \boldsymbol{S})$
    \If {$u \sim \mathrm{U}(0,1) < 0.5$ and $x = S_j$}
        \Comment{Remain longer at stop}
        \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
        \State $\delta \gets \delta - \gamma - d_j$
    \ElsIf {$u \sim \mathrm{U}(0,1) < 0.1$}
        \Comment{Occasionally bus stopped due to traffic}
        \State $w \gets \min(\delta, \mathrm{Exp}(1/\delta))$
        \State $\delta \gets \delta - w$
        \State $z_j \gets z_j + w$
        \Comment{Congestion counts as travel time}
    \EndIf
    \While {$\delta > 0$ and $x < S_J$} 
        \State $\dot x \sim \mathrm{N}(\dot x, \sigma^2)$
        \Comment{System noise, truncated to be non-negative}
        \State $x \leftarrow x + \dot x$
        \If {$x \geq S_{j+1}$}
            \If {$j+1 = J$}
                \Comment{Break at end of route}
                \State $x \gets S_J$
                \Break
            \EndIf
            \State $u_j \sim \mathrm{U}(0,1)$
            \Comment{Simulate vehicle stopping}
            \If {$u_j < \pi$}
                \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
                \State $\delta \gets \delta - \gamma - d_j$
                \State $x \gets S_j$
            \EndIf
            \State $j\gets j+1$
        \EndIf
        \State $\delta \gets \delta - 1$
        \State $z_j \gets z_j + 1$
        \Comment{Track travel time along road}
    \EndWhile
    \\
    \Return $x, \dot x, \boldsymbol z$
    \end{algorithmic}
\end{algorithm}

>>>>

The underlying vehicle state at time $t_k$ consists of
the vehicle's distance travelled $x$ in meters along the route and
its speed $\dot x$ in meters per second.
These are combined into the state vector
$\bX_k = \left[x_k\ \dot x_k\right]^\top$,
of which observations $\bY_k$ are made using a GPS
with GPS error $\omega^2$,
giving the longitude $\lambda_k$ and latitude $\phi_k$ of the vehicle
as $\bY_k = \left[\lambda_k\ \phi_k \right]^\top$.
We use a particle filter to estimate $\bX_k$,
due to its high flexibility and use in recent 
transit vehicle modelling applications \citep{Hans_2015},
as well as many other more general vehicle tracking 
applications \citep{}.
In addition, it makes it trivial to estimate travel times
$\bz = (z_1,\ldots,z_L)^\top$, in seconds, along road segments
as the vehicle traverses the route.

The primary advantage of the particle filter is its handling of multimodality,
as demonstrated in Figure~\ref{fig:pf_state_predict},
which is a common feature of the proposal distribution, particularly around bus stops.
Another advantage is that the likelihood function is intuitively based
on the physical distance between the vehicle observation and the state's
estimate of the true location, $h(\bX_k)$ (Section~\ref{sec:pf_update}).
Conversely, particle filter methods are computationally demanding,
requiring an increasing number of particles as model complexity and
number of parameters increases \citep{Carpenter_1999}.
Section~\ref{sec:rt} describes our implementation of the particle filter
and its \rt performance.

@@

The model components are implemented as an algorithm,
in which (\ref{eq:trans_newton}) is evaluated iteratively
one second at a time, decreasing $\delta_k$ by one each time,
with $\epsilon_k^{(i)}$ only sampled only at the beginning
of each time update.
This allows the particle to approach the next stop,
at which point (\ref{eq:trans_stop}) is evaluated for the particle
and the dwell time subtracted from $\delta_k$,
and continuing until $\delta_k$ reaches zero.
Iterating this way ensures that 
the algorithm works regardless of how many
stops are encountered during the time step.
The algorithm also keeps track of each particle's travel time $z_\ell^{(i)}$
along each segment $\ell$ of the route,
storing the times so that the posterior distribution of the travel time
can be computed once all particles have completed segment $\ell$:
\begin{equation}
\label{eq:post_tt}
p(z_\ell | \bY_{1:k}) \approx
    \sum_{i=1}^N W_k^{(i)} \dot\delta(z_\ell - z_\ell^{(i)}).
\end{equation}

% \renewcommand{\algorithmicrequire}{\textbf{Start:}}
% \newcommand{\algorithmicbreak}{\textbf{break}}
% \newcommand{\Break}{\State \algorithmicbreak}
% \begin{algorithm}[t]
%     \caption{Particle mutation function.}
%     \label{fig:algorithm}
%     \begin{algorithmic}[2]
%     \Require $x_{k-1}, \dot x_{k-1}, \boldsymbol z$\\
%     Initialisation $\delta \gets t_k - t_{k-1}$, 
%         $j \gets stop\_index(x, \boldsymbol{S})$
%     \If {$u \sim \mathrm{U}(0,1) < 0.5$ and $x = S_j$}
%         \Comment{Remain longer at stop}
%         \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
%         \State $\delta \gets \delta - \gamma - d_j$
%     \ElsIf {$u \sim \mathrm{U}(0,1) < 0.1$}
%         \Comment{Occasionally bus stopped due to traffic}
%         \State $w \gets \min(\delta, \mathrm{Exp}(1/\delta))$
%         \State $\delta \gets \delta - w$
%         \State $z_j \gets z_j + w$
%         \Comment{Congestion counts as travel time}
%     \EndIf
%     \While {$\delta > 0$ and $x < S_J$} 
%         \State $\dot x \sim \mathrm{N}(\dot x, \sigma^2)$
%         \Comment{System noise, truncated to be non-negative}
%         \State $x \leftarrow x + \dot x$
%         \If {$x \geq S_{j+1}$}
%             \If {$j+1 = J$}
%                 \Comment{Break at end of route}
%                 \State $x \gets S_J$
%                 \Break
%             \EndIf
%             \State $u_j \sim \mathrm{U}(0,1)$
%             \Comment{Simulate vehicle stopping}
%             \If {$u_j < \pi$}
%                 \State $d_j \sim \mathrm{Exp}(1/\tau_j)$
%                 \State $\delta \gets \delta - \gamma - d_j$
%                 \State $x \gets S_j$
%             \EndIf
%             \State $j\gets j+1$
%         \EndIf
%         \State $\delta \gets \delta - 1$
%         \State $z_j \gets z_j + 1$
%         \Comment{Track travel time along road}
%     \EndWhile
%     \\
%     \Return $x, \dot x, \boldsymbol z$
%     \end{algorithmic}
% \end{algorithm}

% \afterpage{\clearpage}

### 2019-08-08 anzjs2018/sections/04_implementation.tex
more edits

The next most intensive step was vehicle updating,
which took about 5~seconds for 8000 particles.
The ETA prediction step only took a few seconds,
although this may increase once a more comprehensive model is developed.
On our 8-core virtual machine,
the average time to process one iteration during off-peak is under 20~seconds
when 8000~particles are used.
Figure~\ref{fig:day-timings} shows iteration timings over the course of a day.

>>>>

The next most intensive step was vehicle updating,
which took about 5~seconds for 8000 particles.
The ETA prediction step only took a few seconds,
although this may increase once a more comprehensive model is developed.
On our 8-core virtual machine,
the average time to process one iteration during off-peak is about 16~seconds
when 8000~particles are used.
% Figure~\ref{fig:day-timings} shows iteration timings over the course of a day.
